---
title: 'Problem Set #5'
author: "Anaya Hall and Christian Miller"
date: "5/2/2018"
output: pdf_document
fontsize: 11pt
geometry: margin=.75in 
---

```{r setup, include=FALSE}

rm(list = ls())
# Setup
knitr::opts_chunk$set(echo = TRUE, cache = T)
# Options
options(stringsAsFactors = F)
# Packages
library(pacman)
p_load(knitr, kableExtra, tidyverse, dplyr, readr, magrittr, ggplot2, readxl, ascii, sandwich, tinytex)

```

# Part 1: Theory
(Optional -- skip for now!)


# Part 2: Instrumental Variables

## Question 1: NLS80
Revisit the model from *Problem Set #3*, now including ability.

$log(wage) = \beta_0 + exper \cdot \beta_1 + tenure \cdot \beta_2 + married \cdot \beta_3 + south \cdot \beta_4 + urban \cdot \beta_5 + black \cdot \beta_6 + educ \cdot \beta_7 + abil \cdot \gamma + \epsilon$

```{r read_data, message=FALSE}

# Read in CSV as data.frame
wage_df <- readr::read_csv("nls80.csv")

# Select only the variables in our model
wage_df %<>% select(lwage, wage, exper, tenure, married, south, urban, black, educ, iq)
```

### (a) Bias of coefficient on education
Derive the bias of $\beta_7$. Show which direction the bias goes in depending on whether the correlation between ability and education is positive or negative.

$ abil = \delta_0 + exper \cdot \delta_1 + tenure \cdot \delta_2 + married \cdot \delta_3 + south \cdot \delta_4 + urban \cdot \delta_5 + black \cdot \delta_6 + educ \cdot \delta_7 + \eta$


$log(wage) = (\beta_0 + \gamma \delta_0) + (\beta_1 + \gamma \delta_1) \cdot educ + (\beta_2 + \gamma \delta_2) \cdot tenure + (\beta_3 + \gamma \delta_3) \cdot married + (\beta_4 + \gamma \delta_4) \cdot south + (\beta_5 + \gamma \delta_5) \cdot urban  + (\beta_6 + \gamma \delta_6) \cdot black + (\beta_7 + \gamma \delta_7) \cdot educ + \gamma \eta + v$

$plim b_7 = \beta_7 + \gamma \delta_7$

Assume that all $\delta$’s are zero except for the one on the variable of interest (education)

$plim b_7 = \beta_7 + \gamma \cdot \frac {Cov [ abil , educ ]} {Var [educ]}$

Truth is $\beta_7$ , bias is $\gamma \cdot \frac {Cov [ abil , educ ]} {Var [educ]}$

We expect the sign on $\gamma$ to be positive (higher abiltiy should lead to higher wage), the covariance of ability and education to also be positive (more able people acheive higher levels of education), and, of course, the variance of education is positive.
Thus, the bias will also be *positive*.

### (b) Proxy for ability
Estimate the model above excluding ability, record your parameter estimates, standard errors and $R^2$.


First, let's load our OLS function. 
```{r OLS functions}

# Function to convert tibble, data.frame, or tbl_df to matrix
to_matrix <- function(the_df, vars) {
  # Create a matrix from variables in var
  new_mat <- the_df %>%
    #Select the columns given in 'vars'
    select_(.dots = vars) %>%
    # Convert to matrix
    as.matrix()
  # Return 'new_mat'
  return(new_mat)
}


b_ols <- function(y, X) {
  # Calculate beta hat
  beta_hat <- solve(t(X) %*% X) %*% t(X) %*% y
  # Return beta_hat
  return(beta_hat)
}

ols <- function(data, y_data, X_data, intercept = T, H0 = 0, two_tail = T, alpha = 0.05) {
  # Function setup ----
    # Require the 'dplyr' package
    require(dplyr)
  
  # Create dependent and independent variable matrices ----
    # y matrix
    y <- to_matrix (the_df = data, vars = y_data)
    # X matrix
    X <- to_matrix (the_df = data, vars = X_data)
      # If 'intercept' is TRUE, then add a column of ones
      if (intercept == T) {
      X <- cbind(1,X)
      colnames(X) <- c("intercept", X_data)
      }
 
  # Calculate b, y_hat, and residuals ----
    b <- solve(t(X) %*% X) %*% t(X) %*% y
    y_hat <- X %*% b
    e <- y - y_hat
    
  # Useful -----
    n <- nrow(X) # number of observations
    k <- ncol(X) # number of independent variables
    dof <- n - k # degrees of freedom
    i <- rep(1,n) # column of ones for demeaning matrix
    A <- diag(i) - (1 / n) * i %*% t(i) # demeaning matrix
    y_star <- A %*% y # for SST
    X_star <- A %*% X # for SSM
    SST <- drop(t(y_star) %*% y_star)
    SSM <- drop(t(b) %*% t(X_star) %*% X_star %*% b)
    SSR <- drop(t(e) %*% e)
  
  # Measures of fit and estimated variance ----
    R2uc <- drop((t(y_hat) %*% y_hat)/(t(y) %*% y)) # Uncentered R^2
    R2 <- 1 - SSR/SST # Uncentered R^2
    R2adj <- 1 - (n-1)/dof * (1 - R2) # Adjusted R^2
    AIC <- log(SSR/n) + 2*k/n # AIC
    SIC <- log(SSR/n) + k/n*log(n) # SIC
    s2 <- SSR/dof # s^2
  
  # Measures of fit table ----
    mof_table_df <- data.frame(R2uc, R2, R2adj, SIC, AIC, SSR, s2)
    mof_table_col_names <- c("$R^2_\\text{uc}$", "$R^2$",
                             "$R^2_\\text{adj}$",
                             "SIC", "AIC", "SSR", "$s^2$")
    mof_table <-  mof_table_df %>% knitr::kable(
      row.names = F,
      col.names = mof_table_col_names,
      format.args = list(scientific = F, digits = 4),
      booktabs = T,
      escape = F
    )
  
  # t-test----
    # Standard error
    se <- as.vector(sqrt(s2 * diag(solve(t(X) %*% X))))
    # Vector of _t_ statistics
    t_stats <- (b - H0) / se
    # Calculate the p-values
    if (two_tail == T) {
    p_values <- pt(q = abs(t_stats), df = dof, lower.tail = F) * 2
    } else {
      p_values <- pt(q = abs(t_stats), df = dof, lower.tail = F)
    }
    # Do we (fail to) reject?
    reject <- ifelse(p_values < alpha, reject <- "Reject", reject <- "Fail to Reject")
    
    # Nice table (data.frame) of results
    ttest_df <- data.frame(
      # The rows have the coef. names
      effect = rownames(b),
      # Estimated coefficients
      coef = as.vector(b) %>% round(3),
      # Standard errors
      std_error = as.vector(se) %>% round(4),
      # t statistics
      t_stat = as.vector(t_stats) %>% round(3),
      # p-values
      p_value = as.vector(p_values) %>% round(4),
      # reject null?
      significance = as.character(reject)
      )
  
    ttest_table <-  ttest_df %>% knitr::kable(
      col.names = c("", "Coef.", "S.E.", "t Stat", "p-Value", "Decision"),
      booktabs = T,
      format.args = list(scientific = F),
      escape = F,
      caption = "OLS Results"
    )

  # Data frame for exporting for y, y_hat, X, and e vectors ----
    export_df <- data.frame(y, y_hat, e, X) %>% tbl_df()
    colnames(export_df) <- c("y","y_hat","e",colnames(X))
  
  # Return ----
    return(list(n=n, dof=dof, b=b, vars=export_df, R2uc=R2uc,R2=R2,
                R2adj=R2adj, AIC=AIC, SIC=SIC, s2=s2, SST=SST, SSR=SSR,
                mof_table=mof_table, ttest=ttest_table))
}

```
\newpage


```{r model1}
model_1 <- ols(wage_df, y_data = "lwage", 
               X_data = c("exper", "tenure", "married", "south", "urban", "black", "educ"))

model_1$ttest

model_1$mof
```

### (c) Include IQ
(c) Estimate the model including IQ as a proxy, record your parameter estimates, standard errors and $R^2$.

```{r model_iq}
model_iq <- ols(wage_df, y_data = "lwage", 
               X_data = c("exper", "tenure", "married", "south", "urban", "black", "educ", "iq"))

model_iq$ttest

model_iq$mof
```

### (d) Returns on education.
*What happens to returns to schooling? Does this result confirm your suspicion of how ability and schooling
are expected to be correlated?*

Magnitude of the parameter estimate on education decreased. If IQ is a good proxy for ability, this does confirm our suspicion that ability is correlated with education. In the first model, some of the returns on ability (IQ) were mis-attributed to education. In the second model, we correct for this, and see that the parameter estimate on ability is indeed significant.


## Question 2: Recreate results from Card

### (a) Read in data & plot
```{r read_data2, message=FALSE}

# Read in CSV as data.frame
card_df <- readr::read_csv("card.csv")

# Select only the variables in our model
card_df %<>% select(lwage, wage, educ, exper, expersq, black, south, smsa, smsa66, reg661, reg662, reg663, reg664, reg665,reg666, reg667, reg668, nearc4)

head(card_df)
```
``` {r plot_series}
ggplot(data = gather(card_df), aes(x = value)) +
  geom_histogram() +
  facet_wrap(~ key, scales = "free") +
  ggtitle("Histograms of Wage Data variables") +
  ylab("Count") +
  xlab("Value") + theme_minimal()
```

### (b) OLS on log(wage)

``` {r lwage_ols}

rhs_vars <- c("educ", "exper", "expersq", "black", "south", "smsa", "smsa66", "reg661", "reg662", "reg663", "reg664", "reg665", "reg666", "reg667", "reg668")

model1 <- ols(card_df, "lwage", rhs_vars)

model1$ttest

```
*CHECK RESULTS AGAINST PAPER*---> I can't find this ?



### (c) Reduced Form
Estimate reduced form equation for *educ* containing all of the explanatory variables and the dummy variable *nearc4*

```{r reduced_form via OLS}

rhs_vars <- c("nearc4", "exper", "expersq", "black", "south", "smsa", "smsa66", "reg661", "reg662", "reg663", "reg664", "reg665", "reg666", "reg667", "reg668")

rf <- ols(card_df, "educ", rhs_vars)

rf$ttest

```
Yes, the partial correlation between *educ* and *nearc4* IS statistically significant!

### (d) Single IV
Estimate the log(wage) equation by instrumental variables, using nearc4 as an instrument for educ. Compare the 95% confidence interval for the return to educutioan to that obtained from the Least Squares regression above.

### (e) Multiple IV
Use nearc2 and nearc4 as instruments for educ. 
Comment on the significance of the partial correlations of both instruments in the reduced form. 
Show your standard errors from the second stage and compare them to the correct standard errors.

### (f) Hausman test
Conduct a Hausman test for endogeneity of educ. Report your test statistic, critical value and p-value.


# FUNCTIONS FROM PS4

```{r breusch-godfrey test}


# First order only
BGtest <- function(data, y_data, X_data, order = 1) {
  y <- to_matrix(data, y_data)
  X <- to_matrix(data, X_data)
  Z <- X #duplicate rhs matrix
  
  # Run OLS and save residuals to new covariate matrix
  e0 <- ols(data, y_data, X_data)$vars$e
  # Generate 1 time period lagged residuals
  e_lag <- lag(e0)
  e_lag[is.na(e_lag)] <- 0 # Replace NA with zeros

  Z <- cbind(Z, e0)
  
  # Add column of lagged residuals
  Z <- cbind(NA, Z)
  colnames(Z)[1] <- "e_lag"

  # First, convert Z to dataframe for lagging operation
  c <- as.data.frame(Z)
  
  # Create lagged residuals
    for (i in 1:nrow(Z)) {
    if (i == 1)
      c$e_lag[[i]] = 0
    else
      c$e_lag[[i]] = c$e0[i-1]
    }

  # Back to matrix
  X0 <- c[-ncol(c)] %>% as.matrix()
  # BG_df <- cbind(data[y_data], X0)
  BG_df <- c

  # Regress
  # Not regressing on all lhs vars (colnames(X0)), just e_lag
  R2_stat <- ols(BG_df, "e0", "e_lag")$R2
  test_stat <- R2_stat*nrow(data)
  
  pvalue <- 1 - pchisq(test_stat, df = 1)
    
  return(data_frame("Test Statistic" = test_stat, 
              "P-Value" = pvalue))
  
}


# ols(data = gdp_data, y_data = "delta_p", X_data = c("Year", "Realgdp", "Realcons", "Realinvs", "Realgovt", "Realdpi", "CPI_U", "M1", "Tbilrate", "Unemp", "Pop", "Infl", "Realint"))

```


```{r box-pierce test}

BPtest <- function(data, y_data, X_data, order = 1) {
  y <- to_matrix(data, y_data)
  X <- to_matrix(data, X_data)
  Z <- to_matrix(data, X_data)
  
  # Run OLS and save residuals to new covariate matrix
  e0 <- ols(data, y_data, X_data)$vars$e
  Z <- cbind(Z, e0)
  
  # Add column of for lagged residuals
  Z <- cbind(NA, Z)
  colnames(Z)[1] <- "e_lag"

  # First, convert Z to dataframe for lagging operation
  c <- as.data.frame(Z)
  
  # Create lagged residuals
    for (i in 1:nrow(Z)) {
    if (i == 1)
      c$e_lag[[i]] = 0
    else
      c$e_lag[[i]] = c$e0[i-1]
    }

  # Back to matrix
  X0 <- c[-ncol(c)] %>% as.matrix()
  # BG_df <- cbind(data[y_data], X0)
  BP_df <- c

  # Regress e on lagged variables, save coefficient on e_lag
  # Not regressing on all lhs vars (colnames(X0)), just e_lag
  lag_coef <- ols(BP_df, "e0", "e_lag")$b[2]
  test_stat <- nrow(BP_df) * lag_coef^2

  pvalue <- 1 - pchisq(test_stat, df = order)
    
  # return(lag_coef)  
  return(data_frame("Test Statistic" = test_stat,
              "P-Value" = pvalue))
  
  }


```


```{r durbin-watson test}

DWtest <- function(data, y_data, X_data) {
   y <- to_matrix(data, y_data)
  X <- to_matrix(data, X_data)
  Z <- to_matrix(data, X_data)
  
  # Run OLS and save residuals to new covariate matrix
  e0 <- ols(data, y_data, X_data)$vars$e
  Z <- cbind(Z, e0)
  
  # Add column of for lagged residuals
  Z <- cbind(NA, Z)
  colnames(Z)[1] <- "e_lag"

  # First, convert Z to dataframe for lagging operation
  c <- as.data.frame(Z)
  
  # Create lagged residuals
    for (i in 1:nrow(Z)) {
    if (i == 1)
      c$e_lag[[i]] = 0
    else
      c$e_lag[[i]] = c$e0[i-1]
    }

  # # Back to matrix
  # X0 <- c[-ncol(c)] %>% as.matrix()
  # # BG_df <- cbind(data[y_data], X0)
  # DW_df <- c
  # 
  # # Regress e on lagged variables, save coefficient on e_lag
  # lag_coef <- ols(DW_df, "e0", colnames(X0))$b[2]
  
  # numerator summing from t=2
  c1 <- c[-1,]
  d_numer <- sum((c1$e0 - c1$e_lag)^2) 
  d_denom <- sum((c$e0)^2)
  
  # Test statistic
  d <- d_numer/d_denom
  
  return("Test Statistic" = d)
  
}



```


```{r FGLS function, include = FALSE}

fgls <- function(data, y_var, X_vars, Z_vars, intercept = T) {
  # Turn data into matrices
  y <- to_matrix(data, y_var)
  X <- to_matrix(data, X_vars)
  Z <- to_matrix(data, Z_vars)
  # Add intercept
  if (intercept == T) X <- cbind(1, X)
  if (intercept == T) Z <- cbind(1, Z)
  # Calculate n and k for degrees of freedom
  n <- nrow(X)
  k <- ncol(X)
  # Estimate coefficients
  b <- b_ols(y, X)
  # Update names
  if (intercept == T) rownames(b)[1] <- "Intercept"
  # Calculate OLS residuals
  e <- y - X %*% b
  # Regress the squared residuals on Z
  a <- b_ols(e^2, Z)
  # Calculate weights
  w <- Z %*% a
  C <- diag(as.vector(1 / sqrt(w)))
  # Re-weight y and X
  y_tilde <- C %*% y
  X_tilde <- C %*% X
  # Combine the transformed data and run OLS on them
  colnames(X_tilde)[1] <- "Intercept"
  tilde <- cbind(y_tilde, X_tilde) %>% data.frame()
  
  results <- ols(
    data = tilde,
    y_data =  colnames(tilde)[1],
    X_data  = colnames(tilde)[2:ncol(tilde)],
    intercept = F)
  # Return the results
  
  return(results)
}
```


``` {r rho_estimator}
rho_ols <- function(data, y_data, X_data, order = 1) {
  y <- to_matrix(data, y_data)
  X <- to_matrix(data, X_data)
  Z <- to_matrix(data, X_data)
  
  # Run OLS and save residuals to new covariate matrix
  e0 <- ols(data, y_data, X_data)$vars$e
  Z <- cbind(Z, e0)
  
  # Add column of for lagged residuals
  Z <- cbind(NA, Z)
  colnames(Z)[1] <- "e_lag"

  # First, convert Z to dataframe for lagging operation
  c <- as.data.frame(Z)
  
  # Create lagged residuals
    for (i in 1:nrow(Z)) {
    if (i == 1)
      c$e_lag[[i]] = 0
    else
      c$e_lag[[i]] = c$e0[i-1]
    }

  # Back to matrix
  X0 <- c[-ncol(c)] %>% as.matrix()
  # BG_df <- cbind(data[y_data], X0)
  lagged_df <- c

  # Regress e on lagged variables, save coefficient on e_lag 
  # (this is index position 2, after the intercept)
  rho_hat <- ols(lagged_df, "e0", "e_lag")$b[2]
  
  # Alternate way of calculating, gets the same value! 
  # c1 <- c[-1,]
  # rho_numer <- sum((c1$e0 * c1$e_lag)) 
  # rho_denom <- sum((c$e0)^2)
  # rho_1 <- rho_numer/rho_denom
  
  # return(lag_coef)  
  return(rho_hat)
  
}

# Check if function works
# rho_ols(data = gdp_data, y_data = "delta_p", X_data = c("Year", "Realgdp", "Realcons", "Realinvs", "Realgovt", "Realdpi", "CPI_U", "M1", "Tbilrate", "Unemp", "Pop", "Infl", "Realint"))
```


```{r prais-winsten test}

fgls_sc <- function(data, y_var, X_vars, Z_vars, intercept = T, procedure = "prais") {
  # Turn data into matrices
  y <- to_matrix(data, y_var)
  X <- to_matrix(data, X_vars)
  Z <- to_matrix(data, Z_vars)
  # Add intercept
  if (intercept == T) X <- cbind(1, X)
  if (intercept == T) Z <- cbind(1, Z)
  # Calculate n and k for degrees of freedom
  t <- nrow(X)
  k <- ncol(X)
  # Update names
  #if (intercept == T) rownames(b)[1] <- "Intercept"

  # ESTIMATE RHO
  if (procedure == "prais")
      sc_data <- data
  else if (procedure == "cochrane")
      sc_data <- data[-1,]
  
  # Estimate rho for use in FGLS
  rho <- rho_ols(sc_data, y_var, X_vars)
  rho_hat <- rho * (t-k)/(t-1)
  
  # CREATE LOOP THAT BUILDS C MATRIX (is this the right number of rows/columns??)
   G <- matrix(NA, nrow=t, ncol =t)
    for (i in 1:t) {
      for (j in 1:t) {
        if (j == 1 & i == 1) G[i,j] <- sqrt(1-rho)
        else if (i == j) G[i,j] <- 1
        else if (i == j + 1) G[i,j] <- -rho
        else G[i,j] <- 0
      }
    }

  # Re-weight y and X
  y_tilde <- G %*% y
  X_tilde <- G %*% X
  
  # Combine the transformed data and run OLS on them
  colnames(X_tilde)[1] <- "Intercept"
  tilde <- cbind(y_tilde, X_tilde) %>% data.frame()
  
  results <- ols(
    data = tilde,
    y_data =  colnames(tilde)[1],
    X_data  = colnames(tilde)[2:ncol(tilde)],
    intercept = F)
  # Return the results
  
  return(results)
}


```


### (i) Hildreth Lu procedure

```{r Hildreth Lu, eval=F}

rho <- 0.4
data$one <- 1

hildreth_lu <- function(data, y, X1, one, rho) {

#Converting variables to vectors 
y <- select_(data, y) %>% unlist()
X1 <- select_(data, X1) %>% unlist()
one <- select_(data, one) %>% unlist()

#Modifying data based on rho
y_lag <- lag(y)
y_star <- y - rho * y_lag
y_star[1] <- sqrt(1-rho^2) * y[1]


X1_lag <- lag(X1)
X1_star <- X1 - rho * X1_lag
X1_star[1] <- sqrt(1-rho^2) * X1[1]


X2_star <- one - rho
X2_star[1] <- sqrt(1-rho^2)

# Turn data into matrices
  y <- cbind(y_star)
  
  X <- cbind(X1_star, X2_star)
 
#Defining n and k 
n <- nrow(X)
k <- ncol(X) 
  
# Estimate coefficients
  b <- b_ols(y, X)
  
# Calculate OLS residuals
  e <- y - X %*% b
# Calculate s^2
  s2 <- (t(e) %*% e) / (n-k)
  
#Calculate Log likelihood function 
  log_L <- -1*((t(e) %*% e)/ 2*s2) +0.5*log(1-rho^2) -n/2*(log(2*pi) + log(s2))
  
  (log_L)
     
  # Return the results
  return(log_L)
}

#Testing for rho = 0.4 before running loop 
hildreth_lu(data, "y", "unemployment", "one", rho)

#Defining vector of rho's
rhos <- seq(-0.95,0.95, by=0.01)

#Running in a loop 
(LL <- sapply(X = rhos,
             FUN = hildreth_lu,
             data = data, y = "y", X1 = "unemployment",
             one = "one"))

(rhos[which(LL==max(LL))])


```




###### Notes on Parallelizing
library(parallel)
res <- mclapply(query, GET, mc.cores = 4)
map_df(res, function)